Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-07-14 summary.tex
minor edits

\title{Connecting the Popularity Adjusted Block Model to the Generalized Random
Dot Product Graph for Community Detection and Parameter Estimation}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
demonstrating that a PABM is a GRDPG in which communities correspond to
certain mutually orthogonal subspaces of latent positions. This insight
leads to the construction of improved algorithms for community detection
and parameter estimation with PABM. Using established asymptotic
properties of Adjacency Spectral Embedding (ASE) for GRDPG, we derive
asymptotic properties of these algorithms, including algorithms that
rely on Sparse Subspace Clustering (SSC). We illustrate these properties
via simulation.
\end{abstract}

@@

A general family of models that describe unweighted and undirected graphs is the Bernoulli Graph, which is characterized by edge probability matrix $P$. Edges between each pair of vertices $i$ and $j$ are drawn as a Bernoulli trial with probability $P_{ij}$, for each $0 \leq i < j \leq n$ and $n$ is the number of vertices in the graph. Community detection methods often restrict this type of model by assigning each vertex a community label $z_1, ..., z_n$ and then setting each $P_{ij}$ as a function of the labels of vertices $i$ and $j$. Such models are called Block Models.

@@

Parallel to the families of Block Models is another type of Bernoulli Graph 
called the Random Dot Product Graph (RDPG), proposed by
\citet*{10.1007/978-3-540-77004-6_11}. Under this model, each vertex of the graph
can be represented by a point in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where 
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). 
It can be shown that all Bernoulli Graphs with 
positive semidefinite edge probability matrix $P$ is a type of RDPG. 
For instance, the assortative
SBM is equivalent to a special case of the RDPG for which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. A similar property exists for the DCBM
\cite{lyzinski2014, rubindelanchy2017consistency}. 
Because the edge probability matrix for the PABM is not positive semidefinite, 
we instead show that it is a special case of the \emph{Generalized} Random Dot Product Graph (GRDPG) in order to analyze the PABM as a latent space model for community detection and parameter estimation.

@@

Let $G = (V, E)$ be an unweighted, undirected, and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. $A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. We further restrict our analyses to Bernoulli graphs. Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $1 \leq i < j \leq n$  (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). We denote $A \sim \BG(P)$ as graph with adjacency matrix $A$ sampled from edge probability matrix $P$ in this manner. If each vertex has a hidden label in $[K]$, they are denoted as $z_1, ..., z_n$. Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

@@

Another characterization of the PABM is as follows.
Let the rows and columns of $P$ be arranged by community label
such that $n_k \times n_{\ell}$ block $P^{(k \ell)}$
describes the edge probabilities between vertices in communities
$k$ and $\ell$; note that $P^{(k \ell)} = (P^{(k \ell)})^\top$.
Now suppose that each block $P^{(k \ell)}$ can be written as the outer product of two vectors:
\begin{equation} \label{eq:pabm}
  P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top
\end{equation}
for a set of $K^2$ popularity vectors $\{\lambda^{(st)}\}_{1 \leq
  s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}_K$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{Bernoulli}(P)$.
\end{definition}

We will use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\lambda^{(k \ell)}\) consisting of \(K\) underlying communities.

@@

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{Connecting the PABM to the
GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
\begin{proposition}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}
Let
$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix},$$
where the $\{\lambda^{(k \ell)}\}$ are as defined in Definition 1.
Then $A \sim \mathrm{GRDPG}_{3, 1}(X U)$ and 
$B \sim \mathrm{PABM}(\{(\lambda^{(k \ell)}\}_K)$ are
identically distributed.
\end{proposition}

@@

\begin{theorem}[Generalization to $K > 2$]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\})$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by the $\{\lambda^{(k \ell)}\}$ and a $K^2 \times
K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

@@

Let $p = K (K + 1) / 2$ and $q = K (K - 1) / 2$. 
We therefore have
\begin{equation} \label{eq:permutation}
\Pi = U I_{K (K + 1) / 2, K (K - 1) / 2} U^\top
\end{equation}
where $U$ is a $K^2 \times K^2$ orthogonal matrix and hence
\begin{equation} \label{eq:pabm-grdpg}
P = X U I_{p, q} (X U)^\top.
\end{equation}
% \begin{equation} \label{eq:p}
% p = K (K + 1) / 2
% \end{equation}
% \begin{equation} \label{eq:q}
% q = K (K - 1) / 2
% \end{equation}
In summary we can describe the PABM with $K$ communities as a GRDPG with latent
positions $X U$ and signature $(p,q) = \bigl( \tfrac{1}{2} K (K + 1) ,
\tfrac{1}{2} K (K - 1)\bigr)$.
\end{proof}

\begin{example} Let $A$ be a $3$ blocks PABM with latent vectors
  $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. Using the same notation as in Theorem
  \ref{theorem2}, we can define
\begin{gather*}
X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
\end{bmatrix}, \\
Y = \begin{bmatrix}
\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
\end{bmatrix}.
\end{gather*}
Then $Y = X \Pi$ and $P = X Y^{\top}$ where $\Pi$ is a $9 \times 9$ permutation matrix of
the form
$$\Pi = %\begin{bmatrix}
% 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
% 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
% 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
% 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
% 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
% 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
% 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
% 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
% 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1%
\Bigl[e_1 \mid e_4 \mid e_7 \mid e_2 \mid e_5 \mid e_8 \mid e_3
\mid e_6 \mid e_9 \Bigr].
% \end{bmatrix}.
$$
where $e_i$ denote the $i$th basis vector in $\mathbb{R}^{9}$. 
The matrix $\Pi$ corresponds to a permutation of $\{1,2,\dots,9\}$
with the following decomposition.
\begin{enumerate}
\item Positions 1, 5, 9 are fixed.
\item There are three cycles of length 2, namely $(2, 4)$, $(3, 7)$, and $(6, 8)$.
\end{enumerate}
We can therefore write $\Pi$ as $\Pi = U I_{6, 3} U^\top$ where the first three
columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the
{\em fixed} points, the next three columns consist of eigenvectors
$(e_k + e_{\ell}) / \sqrt{2}$, and the last three columns consist of eigenvectors
$(e_k - e_{\ell}) / \sqrt{2}$ for $(k, \ell) \in
\{(2,4),(3,7),(6,8)\}$.

@@

\begin{definition}[Subspace Detection Property]
  \label{def:subspace_detection}
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy
points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the
$y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise
vectors. Let $\lambda \geq 0$ be given and
let $C$ and $B$
be constructed from the
solutions of LASSO problems as described in Eq.~\eqref{eq:ssc} with
this given choice of $\lambda$. Then $X$
is said to satisfy the subspace detection property with weighting parameter $\lambda$ if each column of
$C$ has nonzero norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are
from different subspaces.
\end{definition}

\begin{remark}
In practice, a noisy sample $X$ often does not obey the subspace detection
property. In such cases, $B$ is treated as an affinity matrix for a graph which
is then partitioned into $K$ subgraphs to obtain the clustering. On the other
hand, if $X$ does obey the subspace detection property, $B$ describes a graph
with at least $K$ disconnected subgraphs. Ideally, when the subspace detection
property holds, there are exactly $K$ subgraphs which map to each subspace,
but it could be the case that some of the subspaces are represented by
multiple disconnected subgraphs. The subspace detection property is contingent
on choosing a sufficiently large sparsity parameter $\lambda$.
\end{remark}

@@

If we decompose \(P = Z I_{p, q} Z^\top\), then
\(\exists Q \in \mathbb{O}(p, q)\) such that \(XU = Z Q\). Even if we
start with the exact edge probability matrix, we cannot recover the
``original'' latent positions \(XU\). Note that unlike in the case of
the RDPG, \(Q\) is not necessarily an orthogonal matrix. If \(z_i\)'s
are the rows of \(XU\), then
\(||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2\), and
\(\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle\). This
prevents us from using the properties of \(XU\) directly. In particular,
if \(Q \in \mathbb{O}(n)\), then we could use the fact that
\(\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0\) if
vertices \(i\) and \(j\) are in different communities.

The explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
such that points within each community lie on \(K\)-dimensional
orthogonal subspaces. Multiplication by \(Q \in \mathbb{O}(p, q)\)
removes the orthogonality property but retains the property that each
community is represented by a \(K\)-dimensional subspace. Therefore, the
ASE of \(P\) results in subspaces that correspond to each community,
suggesting the use of SSC. Before exploring SSC, we will first consider
a different approach.

@@

Theorem \ref{theorem3} provides perfect community detection given \(P\).
Letting \(|B|\) be the affinity matrix for graph \(G\), \(G\) is
partitioned into at least \(K\) disjoint subgraphs since each of the
\(K\) communities have no edges between them. Similar to the subspace
detection property, it could be the case that some of the communities
are represented by multiple disjoint subgraphs in \(G\), in which case
additional reconstruction is required to identify the communities
exactly.

Using \(A\) instead of \(P\) introduces error, which converges to \(0\)
almost surely.

@@

Theorems \ref{theorem2}, \ref{theorem3}, and \ref{theorem4} also provide
a natural path toward using SSC for community detection. 
We established in Theorem \ref{theorem2} that an ASE of the edge
probability matrix \(P\) can be constructed such that the communities
lie on mutually orthogonal subspaces, and this property can be recovered
from the eigenvectors of \(P\). Then Theorems \ref{theorem3} and
\ref{theorem4} show that this property holds for the unscaled ASE of
\(A\) drawn from \(P\) as \(n \to \infty\).

@@

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\). Therefore, given
\(P^{(k \ell)}\), \(\lambda^{(k \ell)}\) and \(\lambda^{(\ell k)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, 
let \(P^{(k \ell)} = (\sigma^{(k \ell)})^2 u^{(k \ell)} (v^{(k \ell)})^\top\)
be the singular value decomposition of \(P^{(k \ell)}\).
\(u^{(k \ell)} \in \mathbb{R}^{n_k}\) and 
\(v^{(k \ell)} \in \mathbb{R}^{n_\ell}\) are vectors
and \(\sigma^{(k \ell)}\) is a scalar. 
Then \(\lambda^{(k \ell)} = s_1 u^{(k \ell)}\)
and \(\lambda^{(\ell k)} = s_2 v^{(k \ell)}\) 
for unidentifiable $s_1 s_2 = (\sigma^{(k \ell)})^2$.
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators by taking the SVD of each $A^{(k \ell)}$. 
Since each $\lambda^{(k \ell)}$ is not strictly identifiable,
we instead estimate each $\tilde{\lambda}^{(k \ell)} = \sigma^{(k \ell)} u^{(k \ell)}$.

@@

\[\text{RMSE}(\{\hat{\lambda}^{(k \ell)}\}_K, \{\lambda^{(k \ell)}\}_K) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(k \ell)} - \lambda^{(k \ell)} \|_2^2}\]

@@

Fig \ref{fig:clust_err_ct_sim} shows OSC's community detection error
going to 0 for large \(n\). SSC on both the embedding and on the
adjacency matrix produces similar results for \(K > 2\). Weaker
performance of SSC for \(K = 2\) can be attributed to the final spectral
clustering step of the affinity matrix. 
While the subspace detection property is guaranteed for large \(n\), in our
simulations, setting the sparsity parameter to the required value resulted
in more than \(K\) disconnected subgraphs. 
We instead chose a smaller sparsity parameter, 
necessitating this final clustering step. 
A GMM was fit to the normalized Laplacian eigenmap of $B$,
but visual inspection suggests that the communities are not
distributed as a mixture of Gaussians in the eigenmap. 
A different choice of mixture distribution may result in better performance.

@@

Given ground truth community labels, Algorithm 3 and the MLE-based
plug-in estimators \cite{307cbeb9b1be48299388437423d94bf1} 
perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\).

@@

\caption{Median and IQR RMSE for popularity vectors from Algorithm 3 (red) compared against an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}\label{fig:lambda_est_k}
\end{figure}

We observe similar behavior when comparing errors in the edge probability blocks. 
Both algorithms perform similarly with $n^{-1/2}$ convergence.

>>>>

\title{Connecting the Popularity Adjusted Block Model to the Generalized Random
Dot Product Graph for Community Detection and Parameter Estimation}

\author{John Koo\\Department of Statistics, Indiana University\\
and\\
Minh Tang\\Department of Statistics, North Carolina State University\\
and\\
Michael Trosset\\Department of Statistics, Indiana University}

% \date{\vspace{-2.5em}}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
demonstrating that a PABM is a GRDPG in which communities correspond to
certain mutually orthogonal subspaces of latent positions. This insight
leads to the construction of improved algorithms for community detection
and parameter estimation with PABM. Using established asymptotic
properties of Adjacency Spectral Embedding (ASE) for GRDPG, we derive
asymptotic properties of these algorithms, including algorithms that
rely on Sparse Subspace Clustering (SSC). We illustrate these properties
via simulation.
\end{abstract}

@@

A general family of models that describe unweighted and undirected graphs is 
the Bernoulli Graph, which is characterized by edge probability matrix $P$. 
Edges between each pair of vertices $i$ and $j$ are drawn 
as a Bernoulli trial with probability $P_{ij}$, for each 
$0 \leq i < j \leq n$ and $n$ is the number of vertices in the graph. 
Community detection methods often restrict this type of model 
by assigning each vertex a community label $z_1, ..., z_n$ 
and then setting each $P_{ij}$ as a function of the labels $z_i$ and $z_j$. 
Such models are called Block Models.

@@

Parallel to the families of Block Models is another type of Bernoulli Graph 
called the Random Dot Product Graph (RDPG), proposed by
\citet*{10.1007/978-3-540-77004-6_11}. 
Under this model, each vertex of the graph
can be represented by a vector in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where 
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). 
It can be shown that all Bernoulli Graphs with 
positive semidefinite edge probability matrix $P$ is a type of RDPG. 
For instance, the assortative
SBM is equivalent to a special case of the RDPG for which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. A similar property exists for the DCBM, 
which is equivalent to a RDPG in which vertices that belong to the same 
community all lie in a line segment in the latent space 
\cite{lyzinski2014, rubindelanchy2017consistency}. 
Because the edge probability matrix for the PABM is not positive semidefinite, 
we instead show that it is a special case of the 
\emph{Generalized} Random Dot Product Graph (GRDPG) 
in order to analyze the PABM as a latent space model 
for community detection and parameter estimation.

@@

Let $G = (V, E)$ be an unweighted, undirected, 
and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. 
$A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ 
such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ 
and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. 
We further restrict our analyses to Bernoulli graphs. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. 
Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ 
for each $1 \leq i < j \leq n$ (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). 
We denote $A \sim \BG(P)$ as a graph with adjacency matrix $A$ 
sampled from edge probability matrix $P$ in this manner. 
If each vertex has a hidden label in $[K]$, 
they are denoted as $z_1, ..., z_n$. 
Finally, we denote 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ 
as the collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

@@

Another characterization of the PABM is as follows.
Let the rows and columns of $P$ be arranged by community label
such that $n_k \times n_{\ell}$ block $P^{(k \ell)}$
describes the edge probabilities between vertices in communities
$k$ and $\ell$; note that $P^{(k \ell)} = (P^{(k \ell)})^\top$.
Now suppose that each block $P^{(k \ell)}$ can be written as the outer product of two vectors:

\begin{equation} \label{eq:pabm}
  P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top
\end{equation}

for a set of $K^2$ popularity vectors $\{\lambda^{(st)}\}_{1 \leq
  s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{Bernoulli}(P)$.
\end{definition}

We will use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\{\lambda^{(k \ell)}\}\) consisting of \(K\) underlying communities.

@@

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{Connecting the PABM to the
GRDPG}\label{connecting-the-pabm-to-the-grdpg}}

\begin{proposition}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}
Let

$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix},$$

where the $\{\lambda^{(k \ell)}\}$ are as defined in Definition 1.
Then $A \sim \mathrm{GRDPG}_{3, 1}(X U)$ and 
$B \sim \mathrm{PABM}(\{(\lambda^{(k \ell)}\}_K)$ are
identically distributed.
\end{proposition}

@@

\begin{theorem}[Generalization to $K > 2$]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_K)$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. 
Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by $\{\lambda^{(k \ell)}\}$ and a 
$K^2 \times K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

@@

Let $p = K (K + 1) / 2$ and $q = K (K - 1) / 2$. 
We therefore have

\begin{equation} \label{eq:permutation}
\Pi = U I_{K (K + 1) / 2, K (K - 1) / 2} U^\top
\end{equation}
where $U$ is a $K^2 \times K^2$ orthogonal matrix and hence
\begin{equation} \label{eq:pabm-grdpg}
P = X U I_{p, q} (X U)^\top.
\end{equation}

In summary we can describe the PABM with $K$ communities as a GRDPG with latent
positions $X U$ and signature $(p,q) = \bigl( \tfrac{1}{2} K (K + 1) ,
\tfrac{1}{2} K (K - 1)\bigr)$.
\end{proof}

\begin{example} Let $A$ be a $3$ blocks PABM with latent vectors
  $\{\lambda^{(k \ell)} \colon 1 \leq k \leq 3, 1 \leq \ell \leq 3\}$. Using the same notation as in Theorem
  \ref{theorem2}, we can define
\begin{gather*}
X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & \lambda^{(13)} & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & \lambda^{(21)} & \lambda^{(22)} & \lambda^{(23)} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \lambda^{(31)} & \lambda^{(32)} & \lambda^{(33)}
\end{bmatrix}, \\
Y = \begin{bmatrix}
\lambda^{(11)} & 0 & 0 & \lambda^{(12)} & 0 & 0 & \lambda^{(13)} & 0 & 0 \\
0 & \lambda^{(21)} & 0 & 0 & \lambda^{(22)} & 0 & 0 & \lambda^{(23)} & 0 \\
0 & 0 & \lambda^{(31)} & 0 & 0 & \lambda^{(32)} & 0 & 0 & \lambda^{(33)}
\end{bmatrix}.
\end{gather*}
Then $Y = X \Pi$ and $P = X Y^{\top}$ where $\Pi$ is a $9 \times 9$ 
permutation matrix of the form

$$\Pi = 
\Bigl[e_1 \mid e_4 \mid e_7 \mid e_2 \mid e_5 \mid e_8 \mid e_3
\mid e_6 \mid e_9 \Bigr].$$

where $e_i$ denotes the $i^{th}$ basis vector in $\mathbb{R}^{9}$. 
The matrix $\Pi$ corresponds to a permutation of $\{1,2,\dots,9\}$
with the following decomposition.
\begin{enumerate}
\item Positions 1, 5, 9 are fixed.
\item There are three cycles of length 2, namely $(2, 4)$, $(3, 7)$, and $(6, 8)$.
\end{enumerate}
We can therefore write $\Pi$ as $\Pi = U I_{6, 3} U^\top$ where the first three
columns of $U$ consist of $e_1$, $e_5$, and $e_9$ corresponding to the
{\em fixed} points, the next three columns consist of eigenvectors
$(e_k + e_{\ell}) / \sqrt{2}$, and the last three columns consist of eigenvectors
$(e_k - e_{\ell}) / \sqrt{2}$ for $(k, \ell) \in
\{(2,4),(3,7),(6,8)\}$.

@@

\begin{definition}[Subspace Detection Property]
  \label{def:subspace_detection}
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy
points sampled from $K$ subspaces, i.e., $x_i = y_i + z_i$ where the
$y_i$ belongs to the union of $K$ subspaces and the $z_i$ are noise
vectors. Let $\lambda \geq 0$ be given and
let $C$ and $B$ be constructed from the
solutions of LASSO problems as described in Eq.~\eqref{eq:ssc} with
this given choice of $\lambda$. Then $X$
is said to satisfy the subspace detection property 
with sparsity parameter $\lambda$ if each column of
$C$ has nonzero norm and $B_{ij} = 0$ whenever $y_i$ and $y_j$ are
from different subspaces.
\end{definition}

\begin{remark}
In practice, a noisy sample $X$ often does not obey the subspace detection
property. In such cases, $B$ is treated as an affinity matrix for a graph, 
which requires a partitioning step to obtain the final clustering. On the other
hand, if $X$ does obey the subspace detection property, $B$ describes a graph
with at least $K$ disconnected subgraphs. Ideally, when the subspace detection
property holds, there are exactly $K$ subgraphs which map onto each subspace,
but it could be the case that some of the subspaces are represented by
multiple disconnected subgraphs. The subspace detection property is contingent
on choosing a sufficiently large sparsity parameter $\lambda$.
\end{remark}

@@

If we decompose \(P = Z I_{p, q} Z^\top\), then
\(\exists Q \in \mathbb{O}(p, q)\) such that \(XU = Z Q\). Even if we
start with the exact edge probability matrix, we cannot recover the
``original'' latent positions \(XU\). Note that unlike in the case of
the RDPG, \(Q\) is not necessarily an orthogonal matrix. If \(z_i\)'s
are the rows of \(XU\), then
\(||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2\), and
\(\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle\). This
prevents us from using the orthogonality property of \(XU\) directly.

The explicit form of \(XU\) represents points in \(\mathbb{R}^{K^2}\)
such that points within each community lie on \(K\)-dimensional
orthogonal subspaces. Multiplication by \(Q \in \mathbb{O}(p, q)\)
removes the orthogonality property but retains the property that each
community is represented by a \(K\)-dimensional subspace. Therefore, the
ASE of \(P\) still consists of subspaces 
that correspond to communities of the PABM, even if they are not orthogonal, 
suggesting the use of SSC. Before exploring SSC, we will first consider
a different approach.

@@

Theorem \ref{theorem3} provides perfect community detection given \(P\).
Letting \(|B|\) be the affinity matrix for graph \(G'\), \(G'\) is
partitioned into at least \(K\) disjoint subgraphs since each of the
\(K\) communities have no edges between them. Similar to the subspace
detection property, it could be the case that some of the communities
are represented by multiple disjoint subgraphs in \(G'\), in which case
additional reconstruction is required to identify the communities
exactly.

Using \(A\) instead of \(P\) introduces error, which converges to \(0\)
almost surely. We use this fact to motivate Orthogonal Spectral Clustering 
(Alg. 1).

@@

Theorems \ref{theorem2}, \ref{theorem3}, and \ref{theorem4} also provide
a natural path toward using SSC for community detection. 
We established in Theorem \ref{theorem2} that an ASE of the edge
probability matrix \(P\) can be constructed from a latent vector configuration 
consisting of orthogonal subspaces. Theorem \ref{theorem3} shows how 
this property can be recovered from the eigenvectors of \(P\). 
Then Theorem \ref{theorem4} shows that replacing $P$ with $A$ 
approximates these properties with error that goes to zero as \(n \to \infty\). 
This allows us to obtain the subspace detection property 
in Theorem \ref{theorem5}.

@@

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\). Therefore, given
\(P^{(k \ell)}\), \(\lambda^{(k \ell)}\) and \(\lambda^{(\ell k)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, 
let \(P^{(k \ell)} = (\sigma^{(k \ell)})^2 u^{(k \ell)} (v^{(k \ell)})^\top\)
be the singular value decomposition of \(P^{(k \ell)}\).
\(u^{(k \ell)} \in \mathbb{R}^{n_k}\) and 
\(v^{(k \ell)} \in \mathbb{R}^{n_\ell}\) are vectors
and \(\sigma^{(k \ell)}\) is a scalar. 
Then \(\lambda^{(k \ell)} = s_1 u^{(k \ell)}\)
and \(\lambda^{(\ell k)} = s_2 v^{(k \ell)}\) 
for unidentifiable $s_1 s_2 = (\sigma^{(k \ell)})^2$.
Since each $\lambda^{(k \ell)}$ is not strictly identifiable,
we instead estimate each 
$\tilde{\lambda}^{(k \ell)} = \sigma^{(k \ell)} u^{(k \ell)}$. 
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators by taking the SVD of each $A^{(k \ell)}$ to obtain 
$\hat{\lambda}^{(k \ell)} = \hat{\sigma}^{(k \ell)} \hat{u}^{(k \ell)}$ 
using the largest signular value of $A$ and its corresponding singular vectors.

@@

\[\text{RMSE}(\{\hat{\lambda}^{(k \ell)}\}, \{\tilde{\lambda}^{(k \ell)}\}) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)} \|_2^2}\]

@@

Fig. \ref{fig:clust_err_ct_sim} shows the number of mislabeled vertices
going to 0 for large \(n\). SSC on both the embedding and on the
adjacency matrix produces similar trends for \(K > 2\). Weaker
performance of SSC for \(K = 2\) can be attributed to the final spectral
clustering step of the affinity matrix. 
While the subspace detection property is guaranteed for large \(n\), in our
simulations, setting the sparsity parameter to the required value resulted
in more than \(K\) disconnected subgraphs. 
We instead chose a smaller sparsity parameter, 
necessitating a final clustering step. 
A GMM was fit to the normalized Laplacian eigenmap of $B$,
but visual inspection suggests that the communities are not
distributed as a mixture of Gaussians in the eigenmap. 
A different choice of mixture distribution may result in better performance.

@@

Given ground truth community labels, Algorithm 3 and the MLE-based
plug-in estimators perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\) (Fig. \ref{fig:lambda_est_k}). 
We observe similar behavior when comparing errors 
in the edge probability blocks (Fig. \ref{fig:p_block_est}).

@@

\caption{Median and IQR RMSE for popularity vectors from Algorithm 3 (red) compared against an MLE-based method (blue). Simulations were repeated 50 times for each sample size. Communities were drawn to be approximately balanced.}
\label{fig:lambda_est_k}
\end{figure}
